---
title: "Predicting Spotify Track Commercial Success using Random Forest"
author: "Paweł Kordowski"
date: "2026-01-15"
output:
  pdf_document:
    toc: true            
    toc_depth: 2         
    number_sections: true
    highlight: tango
---

data:https://www.kaggle.com/datasets/maharshipandya/-spotify-tracks-dataset

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```
# Package load
```{r}
library(caret)
library(tidyverse)
library(doParallel)
library(knitr)
library(rsample)
library(vip)
library(pROC)
library(kableExtra)
```
# Data load
```{r}
df <- read_csv("dataset.csv")
```
# Enabling parallel processing
```{r}
n_cores <- detectCores()
cl <- makePSOCKcluster(n_cores)
registerDoParallel(cl)
```
# Dataset breakdown
```{r}
summary(df)
spec(df)
df %>% 
  is.na() %>% 
  colSums()
```
# Popularity threshold
A threshold of 50 was adopted, defining a 'Hit' as a track in the upper quartile of popularity (Top 25%).
```{r}
threshold <- 50
```
# Deleting unnecessary variables and transformation
```{r}
df <- df %>% 
  select(popularity,
         danceability,
         energy,
         loudness,
         speechiness,
         acousticness, 
         instrumentalness, 
         liveness, 
         valence, 
         tempo, 
         duration_ms, 
         explicit, 
         mode, 
         key,
         time_signature) %>%
  mutate(is_popular=ifelse(
    popularity>threshold,
    "Yes",
    "No"
  )) %>% 
  select(-popularity) %>% 
  mutate(
    is_popular=as.factor(is_popular),
    explicit=as.factor(explicit)
)
```
# Train and test split
```{r}
set.seed(42)
train_test_split <- initial_split(
  data = df,
  prop = 0.7
)
df_train <- training(train_test_split)
df_test <- testing(train_test_split)
```
# Model training
```{r}
ctrl <- trainControl(
  method = "cv",
  number = 5, 
  classProbs = TRUE, 
  summaryFunction = twoClassSummary
)

model <- train(
  form = is_popular~.,
  data = df_train,
  method = "rf",
  trControl = ctrl,
  metric = "ROC",
  ntree = 200
)
```
# Predicition
```{r}
prob <- predict(
  object = model,
  newdata = df_test,
  type = "prob"
)[,2]

treshold_rf <- 0.5

pred <- ifelse(
  prob>treshold_rf,
  "Yes",
  "No"
)%>% as.factor()

roc_rf <- roc(
  response = df_test$is_popular,
  predictor = prob
)
plot(roc_rf)
roc_rf$auc

vi(model)
vip(model)
```
The model achieved an AUC score of 0.8552. This indicates that the classifier has a high discriminative ability.

The model correctly identifies most popular tracks, confirming the hypothesis that commercial success largely depends on measurable technical parameters of the track, rather than just random factors.

# Hit profile

The 5 most influential parameters were selected.
```{r}
hit_profile <- df %>% 
  filter(is_popular=="Yes") %>% 
  summarise(
    mean_valence=mean(valence),
    mean_duration=mean(duration_ms)/60000, # to display in minutes
    mean_loudness=mean(loudness),
    mean_acousticness=mean(acousticness),
    mean_tempo=mean(tempo)
)

hit_profile %>% 
  kbl(
    caption = "Hit Profile",
    booktabs = T
  ) %>% 
  kable_styling(
    latex_options = c(
      "striped",
      "hold_position"),
    position = "center"
)
```
The HIT profile reveals that tracks more prone to becoming popular tend to convey sadness or anger (lower valence), last slightly over 3.5 minutes, have a loudness of just under -8 dB, are generally not acoustic, and maintain a tempo of around 121 BPM.
